Sparse spectral and $p$-finite element methods on disk-slices and rhombi I: zero boundary conditions

### 2019-03-04 papers/halfdisk/halfdisk.tex
Some revisions

\begin{abstract}
In recent years, sparse spectral methods have been derived using hierarchies of classical orthogonal polynomials on intervals, disks, and triangles. In this work we extend this methodology to a hierarchy of non-classical orthogonal polynomials on disk slices (e.g. a half disk) and rhombi. This builds on the observation that sparsity is guaranteed due to the boundary being defined by an algebraic curve, and that the entries of partial differential operators can be determined using formulae in terms of (non-classical) univariate orthogonal polynomials. 
\end{abstract}

@@

We show that partial differential equations become sparse linear systems when viewed as acting on expansions involving a family of orthogonal polynomials that  generalise Jacobi polynomials, mirroring the ultraspherical spectral method for ordinary differential equations \cite{SOATUltraspherical} and its analogue on the triangle \cite{SOATGVTriangle}.  On the half-disk the family of weights we consider are of the form
$$
W^{(a,b)}(x,y) = x^a (1-x-y)^b
$$
with corresponding OPs denoted $P_{n,k}^{(a,b)}(x,y)$. Sparsity comes from expanding the domain and range of an operator  using different polynomials. Where the sparsity pattern and entries derived in \cite{SOATGVTriangleRecurrence,SOATGVTriangle}  results from manipulations of Jacobi polynomials, in the present work we use a more general integration-by-parts argument to deduce the sparsity structure, alongside quadrature rules to determine the entries.  Furthermore, we use this framework to derive sparse $p$-finite element methods that are analogous to those of Beuchler and Sch\"oberl on tetrahedra \cite{TriangleFEM}.

The motivation for this work is solving partial differential equations on sub-domains of the sphere. In particular, OPs on a half-sphere can be represented using two families of OPs on the half-disk. Constructing sparse spectral methods for surface PDEs on half-spheres, spherical caps, and spherical triangles is future work, and has applications in weather prediction. Other extensions include a full $hp$-finite element method on sections of a disk, which has applications in fluids.

 We present our procedure  to gain a (two-parameter) family of 2D orthogonal polynomials (OPs) on the half disk domain, by combining 1D OPs to the interval, to form 2D OPs on the disk. We will derive a quadrature rule on the half disk that can be used to expand a function in the OP basis up to a given order. We will demonstrate that these families will have sparse operators (including Jacobi operators representing multiplication by $x$ and $y$, and partial differential operators), and how we can implement function evaluation using the coefficients of the expansion of a given function using the Clenshaw algorithm. We can extend this work directly by projecting up to 3D OPs on the hemisphere, where sparse operators, function evaluation and expansion etc. can be found similarly.

\section{Orthogonal polynomials on the half disk}

@@

The weight $w_2(x)$ is an ultraspherical weight, and the corresponding OPs are   the Jacobi polynomials  \(\{P_n^{(b, b)}\}\). The weight $w_1(x)$ is non-classical and we introduce the notation \(\{H_n^{(a, b)}\}\) for the orthonormal polynomials. Thus we arrive at  a two parameter family of  2D orthogonal polynomials \(\{\Pnk\}\) given by, for \(0 \le k \le n, \: n = 0,1,2,\dots,\)
\begin{align}
 \Pnk^{(a,b)}(x,y) := H_{n-k}^{(a, b+k+\half)}(x) \: \rho(x)^k \: P_k^{(b,b)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align}
orthogonal on \(\Omega\) with respect to the weight $W^{(a,b)}(x,y)$.

@@

Express the three-term recurrences associated to $H_n^{(a,b)}$ and $P_n^{(b,b)}$ as
\begin{align}
x H_n^{(a,b)}(x) &= \beta_n^{(a,b)} H_{n+1}^{(a,b)}(x) + \alpha_n^{(a,b)} H_n^{(a,b)}(x) + \beta_{n-1}^{(a,b)} H_{n-1}^{(a,b)}(x) \\
y P_n^{(b,b)}(y) &= \delta_n^{(b)} P_{n+1}^{(b,b)}(y) + \gamma_n^{(b)} P_n^{(b,b)}(y) + \delta_{n-1}^{(b)} P_{n-1}^{(b,b)}(y),
\end{align}
where we note that \(\gamma_n^{(b)} = 0\) \(\forall n = 0,1,2,\dots\). We can use these to determine the 2D recurences for $P_{n,k}^{(a,b)}(x,y)$:

@@

\begin{proof}
Note that \(\ip<y \Pnkab, P_{m,j}^{(a,b)}>_\Wab = 0\) for \(m < n-1\) (and for \(m > n+2\)). Thus for \(m = n-1, n, n+1,\) \(j = 0,\dots,m:\)
\begin{align}
\ip<y \Pnkab, P_{m,j}^{(a,b)}>_\Wab &= \int \int_\Omega \Pnkab(x,y) \: \Pmjab(x,y) \: y \: \Wab(x,y) \: dy \: dx \\
&= \Big( \int^1_0 H_{n-k}^{(a, b+k+\half)}(s) \: H_{m-j}^{(a, b+j+\half)}(s) \: s^a \: \rho(s)^{2b+k+j+2} \: ds \Big) \nonumber \\
& \quad \quad \quad\cdot \: \Big( \int^1_{-1} P_k^{(b,b)}(t) \: P_j^{(b,b)}(t) \: t \: (1-t^2)^b \: dt \Big) \\
&= \begin{cases}
    	\delta_k^{(b)} \: \ip<H_{n-k}^{(a, b+k+\half)}, H_{m-k-1}^{(a, b+k+\frac{3}{2})}>_{H^{(a, b+k+\frac{3}{2})}} \quad& \text{if } j = k+1 \\
	\delta_{k-1}^{(b)} \: \ip<H_{n-k}^{(a, b+k+\half)}, H_{m-k+1}^{(a, b+k-\half)}>_{H^{(a, b+k+\half)}} \quad& \text{if } j = k-1 \\
	0 & \text{otherwise}
      \end{cases}.
\end{align}

@@

Three-term recurrences lead to Jacobi operators that correspond to multiplication by $x$ and $y$. Define, for some $N \in \No$:
\begin{align}
\bigPab_n := \begin{bmatrix}
		P^{(a,b)}_{n,-n}(x,y) \\
		\vdots \\
		P^{(a,b)}_{n,n}(x,y)
	\end{bmatrix} \in \R^{n+1}, 
\quad \quad 
\bigPab := \begin{bmatrix}
		\bigPab_0 \\
		\bigPab_1 \\
		\vdots \\
		\bigPab_N
	\end{bmatrix}
\end{align}
and set \(J_x^{(a,b)}, J_y^{(a,b)}\) as the \(\half (N+1)(N+2) \times \half (N+1)(N+2)\) Jacobi matrices corresponding to
\begin{align}
J_x^{(a,b)} \bigPab(x,y) = x \bigPab(x,y), \quad J_y^{(a,b)} \bigPab(x,y) = y \bigPab(x,y).
\end{align}
Then  \(J_x^{(a,b)}, J_y^{(a,b)}\) have the Banded-Block-Banded structures
\begin{align}
J_{x/y}^{(a,b)} &= \begin{bmatrix}
		B^{x/y}_0 & A^{x/y}_0 & & & & \\
		C^{x/y}_1 & B^{x/y}_1 & A^{x/y}_1 & & & \\
		& C^{x/y}_2 & B^{x/y}_2 & A^{x/y}_2  & & & \\
		& & C^{x/y}_3 & \ddots & \ddots & \\
		& & & \ddots & \ddots & \ddots \\
	\end{bmatrix}
\end{align}
where
\begin{align}
A^x_n &:= \begin{bmatrix}
		\alphaab_{n+1,0,1} & 0 & \hdots & 0 \\
		& \ddots & & \vdots & \\
		& & \alphaab_{n+1,n,1} & 0 \\
	    \end{bmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots,N-1 \\
B^x_n &:= \begin{bmatrix}
		\alphaab_{n,0,2} & & \\
		& \ddots & \\
		& & \alphaab_{n,n,2} \\
	    \end{bmatrix} \in \R^{(n+1)\times(n+1)} \quad n = 0,1,2,\dots,N \\
C^x_n &:= \big( A^x_n \big)^T \in \R^{(n+1)\times n},  \quad n = 1,2,\dots,N, \\ 
\nonumber \\
A^y_n &:= \begin{bmatrix}
		0 & \betaab_{n,0,6} & & & \\
		\betaab_{n,1,5} & 0 & & & \\
		& \ddots & \ddots & \ddots & \\
		& & \betaab_{n,n,5}& 0 & \betaab_{n,n,6} \\
	    \end{bmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots,N-1 \\
B^y_n &:= \begin{bmatrix}
		0 & \betaab_{n,0,4} & & \\
		\betaab_{n,1,3} & 0 & \ddots & \\
		& \ddots & \ddots & \betaab_{n,n-1,4} \\
		& & \betaab_{n,n,3} & 0
	    \end{bmatrix} \in \R^{(n+1)\times(n+1)}  \quad n = 0,1,2,\dots,N \\
C^y_n &:= \begin{bmatrix}
		0 & \betaab_{n,0,2} & & \\
		\betaab_{n,1,1} & 0 & \ddots & \\
		& \ddots & \ddots & \betaab_{n,n-2,2} \\
		& & \ddots & 0 \\
		& & & \betaab_{n,n,1} \\
	    \end{bmatrix} \in \R^{(n+1)\times n}, \quad n = 1,2,\dots,N.
\end{align}

Note that the sparsity of the Jacobi matrices (in particular the sparsity of the sub-blocks) comes from the natural sparsity of the three-term recurrences of the 1D OPs. In other words, sparsity is not dependent on the domain.

@@

Combining each system in (71) we can write the block-tridiagonal system
\begin{align}
\renewcommand\arraystretch{1.3}
\mleft[
\begin{array}{cccc}
		1 & & & \\
		B_0-G_0(x,y) & A_0 & & \\
		C_1 & B_1-G_1(x,y) & \quad A_1 \quad & \\
		& C_2 & \ddots & \ddots \\
		& & \ddots &
\end{array}
\mright]
\bigPab(x,y)
=
\begin{bmatrix}
	P^{(a,b)}_{0,0} \\ 0 \\ 0 \\ \vdots \\ 0 \\
\end{bmatrix}.
\end{align}
where we note \(P^{(a,b)}_{0,0}(x,y) \equiv H_0^{(a,b)} \: P_0^{(b,b)}\), and for each \(n = 0,1,\dots,N\):
\begin{align}
A_n &:= \begin{bmatrix}
		A^x_n \\
		A^y_n \\
		A^z_n
	    \end{bmatrix} \in \R^{2(n+1)\times(n+3)} \quad (n \ne N), \quad
C_n := \begin{bmatrix}
		C^x_n \\
		C^y_n
	    \end{bmatrix} \in \R^{2(n+1)\times n} \quad (n \ne 0), \nonumber \\
B_n &:= \begin{bmatrix}
		B^x_n \\
		B^y_n
	    \end{bmatrix} \in \R^{2(n+1)\times(n+1)}, \quad
G_n(x,y,z) := \begin{bmatrix}
		xI_{n+1} \\
		yI_{n+1}
	    \end{bmatrix} \in \R^{2(n+1)\times(n+1)}.
\end{align}

For each \(n = 0,1,\dots,N-1\) let \(\Dnt\) be any matrix that is a left inverse of \(A_n\), i.e. such that \(\Dnt A_n = I_{n+3}\). Multiplying our system by the preconditioner matrix that is given by the block diagonal matrix of the \(\Dnt\)'s, we obtain a lower triangular system \citep{dunkl2014orthogonal} (p78), which can be expanded to obtain the recurrence:
\begin{align}
\begin{cases}
\bigPab_{-1}(x,y) := 0 \\
\bigPab_{0}(x,y) := P^{(a,b)}_{0,0} \\
\bigPab_{n+1}(x,y) = -\Dnt (B_n-G_n(x,y)) \bigPab_n - \Dnt C_n  \,\bigPab_{n-1}(x,y), \quad n = 1,2,3,\dots,N.
\end{cases}
\end{align}

Note that we can define \(\Dnt\) as follows. For \(n\) even:
\begin{align}
\Dnt := \begin{bmatrix}
		\frac{1}{\alphaab_{n+1,0,1}} & & &  \\
		& \ddots & & & \\
		& & \frac{1}{\alphaab_{n+1,n,1}} & \\
		0 & \hdots & 0 & \eta_{m-1} & \hdots & 0 & \eta_1 & 0 & \eta_0
	    \end{bmatrix},
\end{align}
where
\begin{align}
m &= \frac{n}{2} + 1, \quad \eta_0 = \frac{1}{\betaab_{n,n,6}}, \nonumber \\
\eta_k &= -\frac{\betaab_{n,n-2(k-1),5} \: \eta_{k-1}}{\betaab_{n,n-2k,6}} \quad k = 1,\dots,m-1.
\end{align}

@@

\begin{align}
\Dnt := \begin{bmatrix}
		\frac{1}{\alphaab_{n+1,0,1}} & &  \\
		& \ddots & & &  \\
		& & \ddots & & \\
		& & & \frac{1}{\alphaab_{n+1,n,1}} & \\
		\xi & 0 & \hdots & 0 & 0 & \eta_{m-2} & \hdots & 0 & \eta_0
	    \end{bmatrix},
\end{align}
 where
\begin{align}
m &= \frac{n+1}{2} + 1, \quad \eta_0 = \frac{1}{\betaab_{n,n,6}}, \nonumber \\
\eta_k &= -\frac{\betaab_{n,n-2(k-1),5} \: \eta_{k-1}}{\betaab_{n,n-2k,6}} \quad k = 1,\dots,m-2, \\
\xi &= -\frac{\betaab_{n,1,5} \: \eta_{m-2}}{\alphaab_{n+1,0,1}}. \nonumber
\end{align}

%
\section{Sparse partial differential operators}

@@

\section{Computational aspects}

@@

%
\section{Examples on the half disk with zero BCs}

>>>>

\begin{abstract}
In recent years, sparse spectral methods for solving partial differential equations have been derived using hierarchies of classical orthogonal polynomials on intervals, disks, and triangles. In this work we extend this methodology to a hierarchy of non-classical orthogonal polynomials on disk slices (e.g. a half disk) and rhombi. This builds on the observation that sparsity is guaranteed due to the boundary being defined by an algebraic curve, and that the entries of partial differential operators can be determined using formulae in terms of (non-classical) univariate orthogonal polynomials. 
\end{abstract}

@@

We show that partial differential equations become sparse linear systems when viewed as acting on expansions involving a family of orthogonal polynomials (OPs) that  generalise Jacobi polynomials, mirroring the ultraspherical spectral method for ordinary differential equations \cite{olver2013fast} and its analogue on the triangle \cite{olver2018recurrence,olver2019triangle}.  On the half-disk the family of weights we consider are of the form
$$
W^{(a,b)}(x,y) = x^a (1-x-y)^b
$$
with corresponding OPs denoted $P_{n,k}^{(a,b)}(x,y)$. We define these to be orthogonalised lexicographically, that is,
$$
P_{n,k}^{(a,b)}(x,y) = C_{n,k} x^{n-k} y^k + (\hbox{lower order terms})
$$
where $C_{n,k} \neq 0$ and ``lower order terms'' can  include degree $n$ polynomials with fewer $y$ terms.

 Sparsity comes from expanding the domain and range of an operator  using different choices of $a$ and $b$. Where the sparsity pattern and entries derived in \cite{olver2018recurrence,olver2019triangle} for equations on the triangle  results from manipulations of Jacobi polynomials, in the present work we use a more general integration-by-parts argument to deduce the sparsity structure, alongside quadrature rules to determine the entries.  Furthermore, we use this framework to derive sparse $p$-finite element methods that are analogous to those of Beuchler and Sch\"oberl on tetrahedra \cite{beuchler2006new}.

The motivation for this work is solving partial differential equations on sub-domains of the sphere. In particular, OPs on a half-sphere can be represented using two families of OPs on the half-disk\sotodo{Explain this}. Constructing sparse spectral methods for surface PDEs on half-spheres, spherical caps, and spherical triangles is future work, and has applications in weather prediction. Other extensions include a full $hp$-finite element method on sections of a disk, which has applications in fluids.

Here is an overview of the paper:

\noindent \secref{OPs}: We present our procedure  to gain a (two-parameter) family of 2D orthogonal polynomials (OPs) on the half disk domain, by combining 1D OPs to the interval, to form 2D OPs on the disk.

\noindent\secref{PDOs}: We demonstrate that these families will lead to sparse operators, including Jacobi operators representing multiplication by $x$ and $y$, and partial differential operators.

\noindent\secref{Computation}: We discuss computational issues, in particular, how to realise the results of the preceding sections in practice.  We  derive a quadrature rule on the half disk that can be used to expand a function in the OP basis up to a given order.  We implement function evaluation using the coefficients of the expansion of a given function using the Clenshaw algorithm.

\noindent\secref{Examples}: We demonstrate the proposed technique for solving Poisson, Helmholtz, and Biharmonic equations on the half-disk.

The motivation for this work is to solve surface PDEs on the sphere.
In subsequent work we will extend  directly by projecting up to 3D OPs on the hemisphere, where sparse operators, function evaluation and expansion etc. can be found similarly.

\section{Orthogonal polynomials on the half disk}\label{Section:OPs}

@@

The weight $w_2(x)$ is an ultraspherical weight, and the corresponding OPs are   the Jacobi polynomials  \(\{P_n^{(b, b)}\}\). The weight $w_1(x)$ is non-classical (it is in fact semi-classical, and is equivalent to a generalized Jacobi weight \cite{magnus1995painleve}) and we introduce the notation \(\{H_n^{(a, b)}\}\) for the orthonormal polynomials. Thus we arrive at  a two parameter family of  2D orthogonal polynomials \(\{\Pnk\}\) given by, for \(0 \le k \le n, \: n = 0,1,2,\dots,\)
\begin{align}
 \Pnk^{(a,b)}(x,y) := H_{n-k}^{(a, b+k+\half)}(x) \: \rho(x)^k \: P_k^{(b,b)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align}
orthogonal on \(\Omega\) with respect to the weight $W^{(a,b)}(x,y)$.

@@

Express the three-term recurrences associated to $H_n^{(a,b)}$ and $P_n^{(b,b)}$ as
\begin{align}
x H_n^{(a,b)}(x) &= \beta_n^{(a,b)} H_{n+1}^{(a,b)}(x) + \alpha_n^{(a,b)} H_n^{(a,b)}(x) + \beta_{n-1}^{(a,b)} H_{n-1}^{(a,b)}(x) \\
y P_n^{(b,b)}(y) &= \delta_n^{(b)} P_{n+1}^{(b,b)}(y) + \gamma_n^{(b)} P_n^{(b,b)}(y) + \delta_{n-1}^{(b)} P_{n-1}^{(b,b)}(y),
\end{align}
where we note that \(\gamma_n^{(b)} = 0\), for \( n = 0,1,2,\dots\). We can use these to determine the 2D recurences for $P_{n,k}^{(a,b)}(x,y)$:

@@

\begin{proof}
Note that \(\ip<y \Pnkab, P_{m,j}^{(a,b)}>_\Wab = 0\) for \(m < n-1\) (and for \(m > n+2\)). Thus for \(m = n-1, n, n+1,\) \(j = 0,\dots,m:\)
\begin{align}
\ip<y \Pnkab, P_{m,j}^{(a,b)}>_\Wab &=  \iint_\Omega \Pnkab(x,y) \: \Pmjab(x,y) \: y \: \Wab(x,y) \: dy \: dx \\
&= \Big( \int^1_0 H_{n-k}^{(a, b+k+\half)}(s) \: H_{m-j}^{(a, b+j+\half)}(s) \: s^a \: \rho(s)^{2b+k+j+2} \: ds \Big) \nonumber \\
& \quad \quad \quad\cdot \: \Big( \int^1_{-1} P_k^{(b,b)}(t) \: P_j^{(b,b)}(t) \: t \: (1-t^2)^b \: dt \Big) \\
&= \begin{cases}
    	\delta_k^{(b)} \: \ip<H_{n-k}^{(a, b+k+\half)}, H_{m-k-1}^{(a, b+k+\frac{3}{2})}>_{H^{(a, b+k+\frac{3}{2})}} \quad& \text{if } j = k+1 \\
	\delta_{k-1}^{(b)} \: \ip<H_{n-k}^{(a, b+k+\half)}, H_{m-k+1}^{(a, b+k-\half)}>_{H^{(a, b+k+\half)}} \quad& \text{if } j = k-1 \\
	0 & \text{otherwise}
      \end{cases}.
\end{align}

@@

Three-term recurrences lead to Jacobi operators that correspond to multiplication by $x$ and $y$. Define, for some $N \in \No$: \sotodo{Why bmatrix, not pmatrix?}
\begin{align}
\bigPab_n := \begin{bmatrix}
		P^{(a,b)}_{n,-n}(x,y) \\
		\vdots \\
		P^{(a,b)}_{n,n}(x,y)
	\end{bmatrix} \in \R^{n+1}, 
\quad \quad 
\bigPab := \begin{bmatrix}
		\bigPab_0 \\
		\bigPab_1 \\
		\vdots \\
		\bigPab_N
	\end{bmatrix}
\end{align}
\sotodo{Why not just make $\bigPab$ infinite for now?}and set \(J_x^{(a,b)}, J_y^{(a,b)}\) as the \(\half (N+1)(N+2) \times \half (N+1)(N+2)\) Jacobi matrices corresponding to
\begin{align}
J_x^{(a,b)} \bigPab(x,y) = x \bigPab(x,y), \quad J_y^{(a,b)} \bigPab(x,y) = y \bigPab(x,y).
\end{align}
Then  \(J_x^{(a,b)}, J_y^{(a,b)}\) have the Banded-Block-Banded structures
\begin{align}
J_{x/y}^{(a,b)} &= \begin{bmatrix}
		B^{x/y}_0 & A^{x/y}_0 & & & & \\
		C^{x/y}_1 & B^{x/y}_1 & A^{x/y}_1 & & & \\
		& C^{x/y}_2 & B^{x/y}_2 & A^{x/y}_2  & & & \\
		& & C^{x/y}_3 & \ddots & \ddots & \\
		& & & \ddots & \ddots & \ddots \\
	\end{bmatrix}
\end{align}
where
\begin{align}
A^x_n &:= \begin{bmatrix}
		\alphaab_{n+1,0,1} & 0 & \hdots & 0 \\
		& \ddots & & \vdots & \\
		& & \alphaab_{n+1,n,1} & 0 \\
	    \end{bmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots,N-1 \\
B^x_n &:= \begin{bmatrix}
		\alphaab_{n,0,2} & & \\
		& \ddots & \\
		& & \alphaab_{n,n,2} \\
	    \end{bmatrix} \in \R^{(n+1)\times(n+1)} \quad n = 0,1,2,\dots,N \\
C^x_n &:= \big( A^x_n \big)^T \in \R^{(n+1)\times n},  \quad n = 1,2,\dots,N, \\ 
\nonumber \\
A^y_n &:= \begin{bmatrix}
		0 & \betaab_{n,0,6} & & & \\
		\betaab_{n,1,5} & 0 & & & \\
		& \ddots & \ddots & \ddots & \\
		& & \betaab_{n,n,5}& 0 & \betaab_{n,n,6} \\
	    \end{bmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots,N-1 \\
B^y_n &:= \begin{bmatrix}
		0 & \betaab_{n,0,4} & & \\
		\betaab_{n,1,3} & 0 & \ddots & \\
		& \ddots & \ddots & \betaab_{n,n-1,4} \\
		& & \betaab_{n,n,3} & 0
	    \end{bmatrix} \in \R^{(n+1)\times(n+1)}  \quad n = 0,1,2,\dots,N \\
C^y_n &:= \begin{bmatrix}
		0 & \betaab_{n,0,2} & & \\
		\betaab_{n,1,1} & 0 & \ddots & \\
		& \ddots & \ddots & \betaab_{n,n-2,2} \\
		& & \ddots & 0 \\
		& & & \betaab_{n,n,1} \\
	    \end{bmatrix} \in \R^{(n+1)\times n}, \quad n = 1,2,\dots,N.
\end{align}

Note that the sparsity of the Jacobi matrices (in particular the sparsity of the sub-blocks) comes from the natural sparsity of the three-term recurrences of the 1D OPs. In other words, sparsity is not dependent on the domain\sotodo{Huh?}.

@@

Combining each system in (71) we can write the block-tridiagonal system
\begin{align}
\renewcommand\arraystretch{1.3}
\mleft[
\begin{array}{cccc}
		1 & & & \\
		B_0-G_0(x,y) & A_0 & & \\
		C_1 & B_1-G_1(x,y) & \quad A_1 \quad & \\
		& C_2 & \ddots & \ddots \\
		& & \ddots &
\end{array}
\mright]
\bigPab(x,y)
=
\begin{bmatrix}
	P^{(a,b)}_{0,0} \\ 0 \\ 0 \\ \vdots \\ 0 \\
\end{bmatrix}.
\end{align}
where we note \(P^{(a,b)}_{0,0}(x,y) \equiv H_0^{(a,b)} \: P_0^{(b,b)}\), and for each \(n = 0,1,\dots,N\):
\begin{align}
A_n &:= \begin{bmatrix}
		A^x_n \\
		A^y_n \\
		A^z_n
	    \end{bmatrix} \in \R^{2(n+1)\times(n+3)} \quad (n \ne N), \quad
C_n := \begin{bmatrix}
		C^x_n \\
		C^y_n
	    \end{bmatrix} \in \R^{2(n+1)\times n} \quad (n \ne 0), \nonumber \\
B_n &:= \begin{bmatrix}
		B^x_n \\
		B^y_n
	    \end{bmatrix} \in \R^{2(n+1)\times(n+1)}, \quad
G_n(x,y) := \begin{bmatrix}
		xI_{n+1} \\
		yI_{n+1}
	    \end{bmatrix} \in \R^{2(n+1)\times(n+1)}.
\end{align}

For each \(n = 0,1,\dots,N-1\) let \(\Dnt\) be any matrix that is a left inverse of \(A_n\), i.e. such that \(\Dnt A_n = I_{n+3}\). Multiplying our system by the preconditioner matrix that is given by the block diagonal matrix of the \(\Dnt\)'s, we obtain a lower triangular system \cite[p78]{dunkl2014orthogonal}, which can be expanded to obtain the recurrence:
\begin{align}
\begin{cases}
\bigPab_{-1}(x,y) := 0 \\
\bigPab_{0}(x,y) := P^{(a,b)}_{0,0} \\
\bigPab_{n+1}(x,y) = -\Dnt (B_n-G_n(x,y)) \bigPab_n - \Dnt C_n  \,\bigPab_{n-1}(x,y), \quad n = 1,2,3,\dots,N.
\end{cases}
\end{align}

Note that we can define an explicit \(\Dnt\) as follows. For \(n\) even:
\begin{align}
\Dnt := \begin{bmatrix}
		\frac{1}{\alphaab_{n+1,0,1}} & & &  \\
		& \ddots & & & \\
		& & \frac{1}{\alphaab_{n+1,n,1}} & \\
		0 & \hdots & 0 & \eta_{m-1} & \hdots & 0 & \eta_1 & 0 & \eta_0
	    \end{bmatrix},
\end{align}
where
\begin{align}
m &= \frac{n}{2} + 1, \quad \eta_0 = \frac{1}{\betaab_{n,n,6}},\qqand \nonumber \\
\eta_k &= -\frac{\betaab_{n,n-2(k-1),5} \: \eta_{k-1}}{\betaab_{n,n-2k,6}} \quad k = 1,\dots,m-1.
\end{align}

@@

\begin{align}
\Dnt := \begin{bmatrix}
		\frac{1}{\alphaab_{n+1,0,1}} & &  \\
		& \ddots & & &  \\
		& & \ddots & & \\
		& & & \frac{1}{\alphaab_{n+1,n,1}} & \\
		\xi & 0 & \hdots & 0 & 0 & \eta_{m-2} & \hdots & 0 & \eta_0
	    \end{bmatrix},
\end{align}
 where
\begin{align}
m &= \frac{n+1}{2} + 1,\quad  \eta_0 = \frac{1}{\betaab_{n,n,6}}, \quad \xi = -\frac{\betaab_{n,1,5} \: \eta_{m-2}}{\alphaab_{n+1,0,1}},\qqand \nonumber \\
\eta_k &= -\frac{\betaab_{n,n-2(k-1),5} \: \eta_{k-1}}{\betaab_{n,n-2k,6}} \quad k = 1,\dots,m-2.
\end{align}
It follows that we can apply $D_n^T$ in $O(n)$ complexity, and thereby construct $\bigPab_{n}(x,y)$ in optimal $O(n^2)$ complexity.

%
\section{Sparse partial differential operators}\label{Section:PDOs}

@@

\section{Computational aspects}\label{Section:Computation}

@@

%
\section{Examples on the half disk with zero BCs}\label{Section:Examples}
